Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-12 manuscript/climate_sensitivity_draft.Rmd
xylem porosity-related changes

Debates have also arisen regarding the traits influencing tree growth responses to drought. It has been observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this classification does not resolve differences among the many species within each category. Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent. For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$) [@scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

@@

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of taller trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height with modest additional affect of crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.  The negative effect of height on $Rt$ held after accounting for species' traits. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

>>>>

Debates have also arisen regarding the traits influencing tree growth responses to drought. *It has been observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this classification does not resolve differences among the many species within each category.* Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses in some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and other forest biomes around the world [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance [@marechaux_leaf_2019], or the direction of response was not always consistent. For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but it correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]. Recent work has shown a great potential for hydraulic traits to predict growth and mortality responses. Hydraulic traits including water potentials at which percent loss of conductivity surpass a certain threshold ($P50$, $P80$, $P88$) and hydraulic safety margin correlate with drought performance [@anderegg_hydraulic_2018] but are time-consuming to measure and therefore infeasible for predicting or modeling drought responses in highly diverse forests (*e.g.*, in the tropics). More easily measurable leaf hydraulic traits with direct linkage to plant hydraulic function can explain greater variation in plant distribution and function [@medeiros_extensive_2019]. These include leaf area shrinkage upon desiccation ($PLA_{dry}$) [@scoffoni_leaf_2014] and the leaf water potential at turgor loss point ($\pi_{tlp}$), *i.e.,* the water potential at which leaf wilting occurs [@bartlett_correlations_2016]. The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree performance under drought remains untested.

@@

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of taller trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height with modest additional affect of crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.  The negative effect of height on $Rt$ held after accounting for species' traits. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].
